Storm surge modelling in the Strait of Georgia

### 2014-12-04 RoughWork/manuscript.tex
Reworked overview of storm surge section

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from the Canadian Tides and Water Level Data Archive \citep{DFOObservations} and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior to the surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

Several significant surge or wind events have been considered in detail, as well as an array of hindcasts to assess model ability in a statistical measure. These cases have been used to assess the model's skill in reproducing water level height and surge amplitude and timing during storm events. The assessment includes comparisons at four permanent tidal measuring stations; Point Atkinson, Victoria, Patricia Bay, and Campbell River (locations indicated in Figure \ref{fig:domain}).  In addition, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variability of the surge as it propagates through the domain and have considered a case with a strong wind event but no significant observed surge.  Finally, comparisons between hindcasts using high resolution and low resolution atmospheric forcing have been completed.

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill.

\subsection{Example Storm Surge Hindcasts}

@@

\subsection{Factors Contributing to Storm Surges}

@@

\subsection{Spatial Extent of Surge}
On Dec 15, 2006, a major storm with wind gusts over 120 km hr$^{-1}$ caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

@@

\subsection{Strong Wind Event}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Climate Data archives \citep{ECClimateArchive}, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. Surprisingly, the strong wind event of Nov 19 was associated with the smaller of the two surges. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

@@

\subsection{High Resolution Atmospheric Forcing}
The above comparison between the modelled and observed winds at Point Atkinson suggests that the atmospheric forcing product (CGRF) may not be adequate to produce accurate storm surge hindcasts since the wind speeds are substantially lower than observed. CGRF is a global atmospheric reforecasting model with horizontal resolution approximately 33 km over the Salish Sea domain, which is problematic since the Strait of Georgia itself is only about 30 km wide. As such, there are only one or two grid cells that are entirely over water in our domain, resulting in overly damped wind speeds at grid points close to land such as Point Atkinson. To determine the effect of under-resolved wind forcing on the storm surge hindcasts we compare simulations that have been forced with 1) the low resolution CGRF model and 2) a high resolution (2.5 km) HRDPS model from Environment Canada.  These simulations are performed for a storm surge that occurred on Dec 17, 2012 and cover the time period Dec 14-19, 2012. The high resolution product is not available for the hindcasts already discussed.

>>>>

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual at four permanent tidal measuring stations; Point Atkinson, Victoria, Patricia Bay, and Campbell River (locations indicated in Figure \ref{fig:domain}). The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from the Canadian Tides and Water Level Data Archive \citep{DFOObservations} and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior to the surge event of interest.

The model's skill is reported through a few statistical measures of the the total water level $\eta$ and the residual $\delta$. First, the mean absolute error ($\bar{e}$), defined as
\begin{equation}
\bar{e} = \text{mean}\left(\left| M - O \right|\right),
\end{equation}
where $M$ and $O$ are the modelled and observed quantities of interest, is a common measure that estimates the average error in the model. Next, a measure comparing the variance of the error to the variance of the observations defined as \citep{thompson2003prediction},
\begin{equation}
\gamma^2 = \frac{\text{var}\left(M-O\right)}{\text{var}\left(O\right)},
\end{equation}
is also calculated. Smaller values of $\gamma^2$ indicate better model skill. This quantity was used by \citet{bernier2006predicting} and \citet{bernier2010tide} in storm surge predictions in Eastern Canada and is calculated here as a point of comparison. A third measure, called the Willmott skill score \citep{willmott1982some}, is also used to assess overall model skill in predicting the water level and residual. This quantity is defined as
\begin{equation}
WS = 1 - \frac{\sum_{i=1}^N \left(M_i - O_i\right)^2}{\sum_{i=1}^N \left(|M_i'| + |O_i'|\right)^2},
\end{equation} 
where $M_i' = M_i-\bar{O}$ and $O_i'=O_i-\bar{O}$ and $\bar{O}$ is the mean of the observations. This score is bounded by zero and one. Values close to one indicate good model skill. These three statisitics, along with the mean observed values and mean modelled valuesm are reported in Table \ref{tab:statistics} for each hindcast. Further, comparisons between the maximum observed and modelled water level and residual are typically reported in the text.

The five hindcast simulations have been used to assess the model's skill and to learn more about how the storm surge propagates through the Salish Sea domain.  First, a general assessment of the model's skill is presented in two hindcasts during Feb 2006 and Nov 2006 in section \ref{sec:hind}. The Feb 2006 case is also used to determine the importance of modelling tidal forcing with the surge propogation. Second, the contribution to the surge due to several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed for the Feb 2006 case in section \ref{sec:factors}. Third, the spatial variability of the surge as it propagates through the domain is examined for a Dec 2006 storm surge in section \ref{sec:spatial}. Fourth, a high wind event that did not result in a large surge or flooding for a Nov 2009 case is discussed in section \ref{sec:wind}. Finally, comparisons between hindcasts using high resolution and low resolution atmospheric forcing is provided in section \ref{sec:res}.

\subsection{Example Storm Surge Hindcasts}\label{sec:hind}

@@

\subsection{Factors Contributing to Storm Surges}\label{sec:factors}

@@

\subsection{Spatial Extent of Surge}\label{sec:spatial}
On Dec 15, 2006, a major storm with wind gusts over 120 km hr$^{-1}$ caused significant damage to many coastal communities on the Pacific Northwest. Although the storm was very strong and the observed residual at Point Atkinson reached 84 cm, this event did not coincide with high tide and so no significant or reported flooding occurred in the Vancouver area \citep{forseth2006adaptation}. Because of the high winds, this event has been used to examine the propagation of the surge throughout the domain and spatial variations in the surge amplitude. First, an assessment of the accuracy of the hindcast is provided.

@@

\subsection{Strong Wind Event}\label{sec:wind}
Next, a relatively strong wind event recorded at Point Atkinson on Nov 19, 2009 is considered. As recorded in the Environment Canada Climate Data archives \citep{ECClimateArchive}, the maximum hourly wind speed at Point Atkinson was 18.6 m s$^{-1}$ and the largest gust observed was 25.8 m s$^{-1}$ or 93 km hr$^{-1}$. In addition, large wind speeds over 15 m s$^{-1}$ persisted for over 9 hours. Yet, the surge associated with this event is rather small when compared to the previous two cases (Figure \ref{fig:nov2009}). Two significant surge events are noted, the first occurring on Nov 19 with an observed amplitude of about 50 cm at Point Atkinson, and a second, larger surge of 62.5 cm on Nov 20. Surprisingly, the strong wind event of Nov 19 was associated with the smaller of the two surges. Neither of the surges occurred during a high tide and so no high water level events were recorded. Although not shown here, the other three locations also observed a maximum surge on Nov 20, which appears to be linked to a secondary high wind event and another surge entering the system at Tofino and Port Hardy.

@@

\subsection{High Resolution Atmospheric Forcing}\label{sec:res}
The above comparison between the modelled and observed winds at Point Atkinson suggests that the atmospheric forcing product (CGRF) may not be adequate to produce accurate storm surge hindcasts since the wind speeds are substantially lower than observed. CGRF is a global atmospheric reforecasting model with horizontal resolution approximately 33 km over the Salish Sea domain, which is problematic since the Strait of Georgia itself is only about 30 km wide. As such, there are only one or two grid cells that are entirely over water in our domain, resulting in overly damped wind speeds at grid points close to land such as Point Atkinson. To determine the effect of under-resolved wind forcing on the storm surge hindcasts we compare simulations that have been forced with 1) the low resolution CGRF model and 2) a high resolution (2.5 km) HRDPS model from Environment Canada.  These simulations are performed for a storm surge that occurred on Dec 17, 2012 and cover the time period Dec 14-19, 2012. The high resolution product is not available for the hindcasts already discussed.
